---
title: 'Looking for a reset button: How Jim Boylen changed the 2019 Bulls'
author: Sean Boren
date: '2019-01-24'
slug: looking-for-a-reset-button-how-jim-boylen-changed-the-2019-bulls
categories:
  - BULLS
tags: []
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, error = FALSE, warning = FALSE, message = FALSE)

```

Since taking over for Fred Hoiberg at the beginning of December, Jim Boylen has taken a ton of 
criticism for his approach as the new lead man for the Bulls, but rightfully so. His "play hard" 
mentality and push to be the best conditioned team in the league (defending push-up champs) hasn't 
resonated with players of this era. Not only is this management style not fitting well, his 
philosophy as a coach doesn't fit the mold of how the league has trended with a barrage of three 
point attempts coming per game. While this is rather easy to see now after 25 games, I initially 
wanted to reserve any judgement on the new coach and see how the team would perform and if the 
young core would develop. Well, after a 20 point drubbing from the rebuilding Atlanta Hawks at home, 
I think it's safe to say the team has some work to do. Not only do we not see any cohesiveness from 
the team on the court, but we are starting to sense some divide amongst the players and the direction 
their coach is taking them.

```{r}
blogdown::shortcode('tweet', '1088602351747977218')
```

Now that Boylen has at least matched the number of games Hoiberg has coached this year, let's see how the 
team has transformed. 

### Shot Selection

There's a reason we've seen the three point shot become the focal point for most teams: analytics. 
It really is simple math, but the evolution of analytics in sports has made front offices realize 
that attempting more three-pointers will increase your expected point total. Well, since drill 
sergeant Boylen took over, the Bulls have consistently taken less threes. 

```{r results = 'hide'}
library(tidyverse)
library(zoo)
library(nbastatR)
library(ggimage)
library(rsvg)

team_data <- game_logs(seasons = 2019, result_types = "team")

bulls_data <- filter(team_data, nameTeam == "Chicago Bulls") %>%
  mutate(head_coach = ifelse(numberGameTeamSeason < 25, "Fred Hoiberg", "Jim Boylen"))

```

```{r fig.align='center'}
league_average <- mean(sum(team_data$fg3aTeam)/sum(team_data$fgaTeam))

# 5 game rolling proportion of 3 point shots
bulls_data %>%
  mutate(prop_threes = fg3aTeam/fgaTeam,
         roll_sum_3att = rollsum(fg3aTeam, 5, align = "right", fill = NA),
         roll_sum_fga = rollsum(fgaTeam, 5, align = "right", fill = NA),
         prop_threes_roll = roll_sum_3att/roll_sum_fga) %>%
  ggplot(aes(x = dateGame, y = prop_threes_roll)) + 
    geom_point(color = "#CE1141") +
    geom_smooth(method = "loess", se = FALSE, span = 0.25, color = "gray", size = 0.2) + 
    #geom_vline(xintercept = as.numeric(as.Date("2018-12-08")), linetype = "dashed") +
    geom_hline(yintercept = league_average, linetype = "dotted") +
    scale_y_continuous(breaks = seq(0, 1, 0.02), labels = function(x) paste0(x*100, "%")) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_line(color = "gray85"),
          panel.background = element_rect(fill = "gray95"),
          plot.background = element_rect(fill = "gray95"),
          axis.ticks = element_blank(),
          axis.text = element_text(size = 12),
          axis.title.y = element_text(margin = margin(0, 0.5, 0, 0, "cm"), size = 14), 
          plot.title = element_text(size = 20, margin = margin(0,0,0.5,0, "cm")),
          plot.subtitle = element_text(size = 14, vjust = 4),
          plot.margin = margin(0.75,0.75,0.75,0.75,"cm")) +
    #annotate("text", x = as.Date("2018-11-10"), y = 0.385, label = "Hoiberg", fontface = "bold", size = 6) +
    #annotate("text", x = as.Date("2019-01-07"), y = 0.385, label = "Boylen", fontface = "bold", size = 6) +
    annotate("text", x = as.Date("2018-12-25"), y = league_average + 0.003, 
             label = paste0("League Average ", round(league_average*100), "%")) +
    labs(title = "The making of an old school offense",
         subtitle = "% of FG that are 3PA (5-game moving average)",
         x = NULL,
         y = "3PA Rate",
         caption = "ChiSports Data   |   source: NBA.com via nbastatR")
```

On average across the league, the 3-point shot represents 35% of teams’ field goal attempts. The 
Bulls are consistently below that when looking across a 5-game average and hit peak “Boyball” at the 
end of December into early January. Only recently are we seeing the team begin to take more 3-pointers 
which is likely due to a little talking to from the front office.

### Pace

We've read ad nauseam and seen just how much Jim Boylen wanted to slow the pace down for this team.
The team has recently been upping the pace a bit and utilizing it effectively to get points
in transition, but that wasn't the case in Boylen's earlier games. The reason so many have called 
for the Bulls to play uptempo is because of the makeup of this roster. With such a young group of 
players, the coach should want to try and run the other team out of the gym. The Atlanta Hawks are
currently ranked 1st in pace and while their win percentage won't reflect it, the team has found a 
way to make their players successful by getting up and down the floor. If Boylen really wants to try 
and play in the half-court setting, though, its not that you can't be successful doing that. The 
relationship between pace and a team's win percentage isn't linear. 

```{r results = 'hide'}
team_stats <- teams_players_stats(seasons = 2019, types = "team", tables = "general", 
                                  season_types = "Regular Season", measures = "Advanced", modes = "PerGame")

```

```{r}

team_stats$dataTable[[1]] %>%
  inner_join(select(team_data, nameTeam, urlTeamSeasonLogo), by = "nameTeam") %>%
  ggplot(aes(x = paceRank, pctWins)) + 
  geom_image(aes(image = urlTeamSeasonLogo), size = 0.1) + 
  scale_size_identity() +
  scale_y_continuous(breaks = seq(0,1,0.10)) +
  theme(panel.grid = element_line(color = "gray85"),
        panel.background = element_rect(fill = "gray95"),
        plot.background = element_rect(fill = "gray95"),
        axis.ticks = element_blank(),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        axis.title.y = element_text(margin = margin(0,0.5,0,0, "cm")),
        axis.title.x = element_text(margin = margin(0.5,0,0,0, "cm")),
        plot.title = element_text(size = 22, margin = margin(0,0,0.5,0, "cm"), vjust = 5, face = "bold"),
        plot.margin = margin(0.75,0.75,0.75,0.75,"cm"),
        plot.caption = element_text(margin = margin(0.5,0,0,0, "cm"))) +
  labs(x = "Pace (League Rank)",
       y = "Win Percentage",
       title = "Pace alone doesn't lead to more wins",
       caption = "ChiSports Data   |   source: NBA.com via nbastatR")

```

The Spurs, Nuggets, and Rockets are all in the bottom 6 of the league when it comes to pace, 
possessions per 48. Unlike the Bulls, though, these teams employ a very active, motion offense in the 
half-court. Their offensive sets have allowed the Spurs to have the best 3P FG% in the league, while 
the Nuggets have the 2nd best assist ratio (assists per 100 poss.) in the league. The Rockets on the 
other hand, well, they have James Harden who is having a record-setting year. 

Slow pace **can** work, but its how you utilize the offense to generate open shots - something the Bulls 
haven't quite done. In watching games, the team doesn't look to have a real plan. There is far too much
put on the shoulders of Zach Lavine to make plays and Lauri Markkanen to attempt a low percentage 
shot at the end of a shot clock. This team, and mainly its coach, need to make adjustments and 
find what can help make them successful. In the case of a rebuilding organization, successful doesn't 
have to be defined by wins. At the very least it would mean remaining competitive in most 
games which is something they were doing under Fred Hoiberg. Just like the players, I'm sick of seeing 
blowout losses and something needs to change.
