---
title: Evaluating Joc Pederson
author: Sean Boren
date: '2019-01-21'
slug: evaluating-joc-pederson
categories:
  - White Sox
tags: []
draft: TRUE
---
```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, error = FALSE, warning = FALSE, message = FALSE)

```

With rumblings about a potential trade for Joc Pederson, the White Sox are continuing to remain 
active this offseason, although we are all waiting for that one **BIG** signing to occur still. If 
you're active on Twitter, you'll notice a lot of your favorite White Sox accounts would tell you 
that Pederson would have ranked 1st in almost every statistical category for the White Sox last 
season. Let's take a look at some of the numbers and see if we can pinpoint what Joc Pederson brings
to the table for the South Siders. 

# What we know

```{r}
library(tidyverse)
library(baseballr)

joc_id <- playerid_lookup("Pederson", "Joc")
```

Joc made his major league debut at the end of 2014 for the Dodgers at age 22 and although he for a
0.210 average in 2015, the outfielder was selected to his first All-Star game. 

```{r data-import}
fg_hit_data <- fg_bat_leaders(2015, 2018, qual = 400*4, ind = 0)

joc_data <- filter(fg_hit_data, Name == "Joc Pederson")

fg_2018 <- fg_bat_leaders(2018, 2018, qual = 400)
```

# The Good 

Of `r nrow(fg_hit_data)` qualified hitters that averaged over 400 plate appearances since 2015, 
Pederson has a wOBA better than `r sum(joc_data$wOBA > fg_hit_data$wOBA)` players, or about
`r scales::percent(round(mean(joc_data$wOBA > fg_hit_data$wOBA), 2))` of the field. 

```{r woba-plot}
ggplot(fg_hit_data, aes(x = wOBA)) + 
  geom_histogram() +
  geom_vline(xintercept = joc_data$wOBA, color = "dodgerblue", linetype = "dashed") +
  annotate("text", x = joc_data$wOBA - 0.002, y = 7, 
           label = "Pederson's wOBA",
           angle = 90, size = 4, color = "white") +
  annotate("text", x = max(fg_hit_data$wOBA), y = 2,
           label = "Mike Trout", size = 4) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 25)) +
  theme(plot.background = element_rect(fill = "gray95"), 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_line(color = "gray85"),
        panel.background = element_rect(fill = "gray95"),
        axis.ticks = element_blank(),
        axis.text = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, face = "bold", margin = margin(b = 30)),
        plot.subtitle = element_text(size = 12, vjust = 10)) +
  labs(x = "wOBA",
       y = NULL,
       title = "Pederson's wOBA MLB rank",
       subtitle = "hitters with at lease 1,600 PA over last 4 seasons",
       caption = "source: Fangraphs | baseballr")
```

As others have mentioned, he's ranked 1st in most categories if he were part of the forgettable 2018
White Sox. His `r filter(fg_2018, Name == "Joc Pederson")$wOBA` wOBA in 2018 is a fair number better
than the best performing Sox players. Of those with at least 400 plate appearances in 2018, here's 
how they faired. If this were Twitter, I'd have to add a #108ing #Palkasmash.

```{r}
fg_2018 %>% 
  filter(Team == "White Sox") %>% 
  arrange(desc(wOBA)) %>% 
  select(Name, wOBA) %>%
  knitr::kable()
```

In addition to wOBA, here's a list of categories Joc would've ranked first in for the 2018 White Sox: 

  * WAR
  * OPS
  * SLG
  * Walk to Strikeout Ratio
  * ISO
  * K Rate

```{r, eval = FALSE}
fg_2018 %>%
  filter(Team == "White Sox" | Name == "Joc Pederson") %>%
  select(Name, H:WPA) %>%
  gather(key = metric, value = value, -Name) %>%
  group_by(metric) %>%
  arrange(desc(value)) %>%
  mutate(rank = row_number()) %>%
  filter(Name == "Joc Pederson" & rank == 1)
```






